Integrating Common Ground and Informativeness in Pragmatic Word Learning

### 2019-02-05 papers/cogsci2019/cogsci2019.Rmd
update cogsci submission

```{r, libraries}
library(png)
library(grid)
library(ggplot2)
library(xtable)
library(tidyverse)
library(knitr)
library(ggthemes)
library(langcog)
library(rwebppl)
library(matrixStats)
library(coda)
library(ggpubr)
library(lme4)
library(broom)
library(exactRankTests)
```

```{r}
# data ex 1 and ex 2
d1 <- bind_rows(read_csv(file="../../stats/data/novel.data.csv"),
               read_csv(file="../../stats/data/pref.data.csv")) %>%
  mutate(trial_type = ifelse(trial == "train", "train", "test"),
         control = ifelse(change == "true", "true" , "false"))

d2 <- read_csv(file="../../stats/data/inf.data.csv") %>%
  mutate(trial_type = ifelse(trial == "train", "train", "test"))

@@

pred_ex4 <- readRDS(file = "stats/saves/model_predictions_ex3_variable_priors.rds")

bf_ex3 <- readRDS(file = "stats/saves/ex3_bf.rds")

bf_ex4 <- readRDS(file = "stats/saves/ex4_bf.rds")

@@

Within RSA models, each agent in this recursion is modeled as a Bayesian reasoner; thus, information integration is treated as a process of probabilistic inference. The speaker-general informativeness expectation is already encoded in the structure of the model: Speakers produce utterances to aid the listener in disambiguating referents. We operationalize speaker-specific, common ground information as the shared prior probability of referents in the context of the utterance. Thus, a natural locus for information integration within these models is the trade off between the prior probability of a particular referent and the likelihood of that referent given the current utterance.

Here we evaluate this rational, pragmatic account of information integration. We isolate speaker-specific and common-ground information experimentally, then test how adult listeners combine them in a word learning setting. In Experiments 1 and 2, we replicate findings showing that listeners expect speakers to a) produce informative utterances (Experiment 1) and b) communicate about things that are relevant to common ground (Experiment 2). Based on these results, we generate model predictions using the RSA framework about how these two components should be integrated. In Experiment 3, we test how listeners integrate their expectations and compare model predictions to empirical data. Experiment 4 replicates and extends Experiment 3 by varying the strength of common ground assumptions.

@@

For all experiments, we pre-registered the sample size, experimental design and the statistical analysis. For Experiment 3 and 4, we also registered the model structure and predictions (see [masked for peer review])

```{r design, fig.env = "figure*", fig.pos = "h", fig.width=6, fig.height=5, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "Schematic experimental procedure. In all conditions, at test (bottom), the speaker ambiguously requested an object using a non-word (e.g. “dax”). Participants clicked on the object they thought the speaker referred to. Informativeness (Experiment 1, left) translated to making one object less frequent in context. Common ground (Experiment 2, middle) was manipulated by making one object prefered by or new to the speaker. Green plus signs represent an expression of preference and red minus of dispreference (see main text for details). When expressing e.g. preference for an object on a table with two objects (panel 3 from top), the respective object was temporarily enlarged. Experiment 3 (right) combined manipulations. Condition shown here: preference - same speaker - incongruent."}
img <- png::readPNG("figs/method_figure.png")
grid::grid.raster(img)
```

@@

A total of 121 individuals participated in the experiment. The test situation was the same as in the test condition in Experiment 1: One table with object of type A and the other with an object of type A and B. Again, the animal always turned to the table with two objects and ambiguously requested an object. In Experiment 1, the listener had no prior information about each object. In Experiment 3, however, we manipulated common ground expectations in the same way as in Experiment 2. For example, the animal would turn to the table with one object and express that they don’t like object A, then turn to the other table and express that they like object B. Next, after quickly disappearing, they would reappear, turn to the table with two objects and make a request (see Fig. \ref{fig:design}, right).

For each common ground condition, there were 4 conditions in Experiment 3 resulting from the cross of congruent/incongruent informativeness with same/different speaker. If the preferred/novel object was the less frequent one (object B), the two expectations were congruent. If the preferred/novel object was the more frequent one (object A), expectations were incongruent. For each type of expectation alignment, we varied if the same or a different animal returned. Participants either completed the preference or novelty version with two test trials in each of the four conditions. Before discussing the empirical results, we briefly discuss the cognitive model we used to predict expectation integration.

@@

To derive predictions, we used a probabilistic RSA model that simulates a pragmatic listener reasoning about a cooperative speaker who is trying to refer to an object [@frank2012predicting]. The speaker chooses how to refer to the object by reasoning about a naive listener who does not know the labels for the object [@frank2014inferring]. The conditional probability that the listener chooses a referent given an utterance is defined as follows:
$$P_L(r_s|u)\propto P_S(u|r_s)P_S(r_s)$$
Here, $P_S(u|r_s)$ is the likelihood that the speaker will use an utterance u to refer to a specific referent r. It is defined in terms of a utility function $U_S(u;s)$ consisting of the surprisal of $u$ for a naive listener $L_0$, who interprets $u$ according its literal semantics:
$$P_S(u|r_s)\propto exp(\alpha U_S(u;s))$$
The numerical strength of the expression above depends on a scalar value, $\alpha$, which can be interpreted as an indicator of how rational the speaker is in choosing utterances (i.e. as $\alpha$ increases, the speaker is more likely to choose the most informative utterance).

@@

Results are discussed in the form of the proportion with which listeners chose the more informative object (i.e., the object that would be the more informative referent when only considering speaker general information). For a comparison to chance within each unique condition see \ref{fig:plotexp3}. Combinations of alignment and speaker identity differed in how they influenced participants’ responses (GLMM model term: `alignment*speaker`; $\beta$ = `r lm_ex3_r%>%filter(term == "SpeakerSame speaker:AlignmentIncongruent")%>%pull(estimate)`, se = `r lm_ex3_r%>%filter(term == "SpeakerSame speaker:AlignmentIncongruent")%>%pull(std.error)`, *p* `r lm_ex3_r%>%filter(term == "SpeakerSame speaker:AlignmentIncongruent")%>%pull(p.value)`). Fig. \ref{fig:plotmodelcomp}A and B shows the mean response in each unique condition compared to the pragmatics model. Model predictions and data were highly correlated (r = `r model_comp_r$estimate`, *p* `r model_comp_r$p.value`).Model fit was much better in the model taking into account both types of expectations compared to the uniform prior (BF = `r format(bf_ex3%>%filter(Comparison == "Pragmatic vs. No prior") %>%pull(BayesFactor),digits = 2)`) or prior only model (BF = `r format(bf_ex3%>%filter(Comparison == "Pragmatic vs. Prior only") %>%pull(BayesFactor), digits = 2)`). The inferred noise level in the pragmatics model was 0.27 (95% HDI: 0.21 - 0.34].

@@

As noted above, the strong prior condition was a direct replication of Experiment 3. Results from the two rounds of data collection were highly correlated (r = `r repl_r$estimate`, *p* `r repl_r$p.value`, see Fig. \ref{fig:plotmodelcomp}C). Across levels of prior manipulation, the data from Experiment 4 were highly correlated to the corresponding model predictions (r = `r ex4_comp_r$estimate`, *p* `r ex4_comp_r$p.value`, see Fig. \ref{fig:plotmodelcomp}B). Again, the pragmatics model provided a much better fit compared to the flat prior (BF = `r format(bf_ex4%>%filter(Comparison == "Pragmatic vs. No prior") %>%pull(BayesFactor),digits = 2)`) or prior only model (BF = `r format(bf_ex4%>%filter(Comparison == "Pragmatic vs. Prior only") %>%pull(BayesFactor), digits = 2)`). The inferred noise level in the pragmatics model was 0.28 (95% HDI: 0.24 - 0.32].

@@

Language use and learning requires balancing different types of expectations about one’s interlocutor - both expectations about how speakers behave in general and specific expectations about how a particular speaker might behave in a particular context. Here we used a Bayesian pragmatics model to predict this integration process. Experiment 1 and 2 replicated previous studies showing that adult listeners expect speakers to produce utterances informatively and also with respect to common ground. We then combined the procedures from the first two experiments to study how listeners would integrate expectations. We used the results from Experiment 1 and 2 to specify model parameters that represented the two types of expectations, generating predictions about new behavior. Experiments 3 and 4 showed that both types of expectations influenced listeners inferences. Overall, listener behavior was accurately described by our model, suggesting that listeners trade-off flexibly between speaker specific and general pragmatic expectations.

>>>>

```{r, libraries}
library(png)
library(grid)
library(ggplot2)
library(xtable)
library(tidyverse)
library(knitr)
library(ggthemes)
library(langcog)
library(matrixStats)
library(coda)
library(ggpubr)
library(lme4)
library(broom)
library(exactRankTests)
```

```{r}
# data ex 1 and ex 2
  d1 <- bind_rows(read_csv(file="../../stats/data/novel.data.csv"),
               read_csv(file="../../stats/data/pref.data.csv")) %>%
  mutate(trial_type = ifelse(trial == "train", "train", "test"),
         control = ifelse(change == TRUE, "true" , "false"))

d2 <- read_csv(file="../../stats/data/inf.data.csv") %>%
  mutate(trial_type = ifelse(trial == "train", "train", "test"),
         control = ifelse(control == TRUE, "true", "false"))

@@

pred_ex4 <- readRDS(file = "saves/model_predictions_ex3_variable_priors.rds")

bf_ex3 <- readRDS(file = "saves/ex3_bf.rds")

bf_ex4 <- readRDS(file = "saves/ex4_bf.rds")

@@

Within RSA models, each agent in the recursion is modeled as a Bayesian reasoner; thus, information integration is treated as a process of probabilistic inference. The speaker-general informativeness expectation is already encoded in the structure of the model: Speakers produce utterances to aid the listener in disambiguating referents. We operationalize speaker-specific, common ground information as the shared prior probability of referents in the context of the utterance. Thus, a natural locus for information integration within these models is the trade off between the prior probability of a particular referent and the likelihood of that referent given the current utterance.

Here we evaluate this rational, pragmatic account of information integration. We isolate speaker-specific and common-ground information experimentally, then test how adult listeners combine them in a word learning setting. In Experiments 1 and 2, we replicate findings showing that listeners expect speakers to a) produce informative utterances (Experiment 1) and b) communicate about things that are relevant to common ground (Experiment 2). Based on these results, we generate model predictions using the RSA framework about how these two components should be integrated. In Experiment 3, we test how listeners integrate their expectations and compare model predictions to empirical data. Experiment 4 replicates and extends Experiment 3 by varying the strength of common ground assumptions. For all experiments, we pre-registered the sample size, experimental design and the statistical analysis. For Experiment 3 and 4, we also registered the model structure and predictions (see [masked for peer review])

@@

```{r design, fig.env = "figure*", fig.pos = "h", fig.width=6, fig.height=5, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "Schematic experimental procedure. In all conditions, at test (bottom), the speaker ambiguously requested an object using a non-word (e.g. “dax”). Participants clicked on the object they thought the speaker referred to. Informativeness (Experiment 1, left) translated to making one object less frequent in context. Common ground (Experiment 2, middle) was manipulated by making one object prefered by or new to the speaker. Green plus signs represent utterances that expressed preference and red minus of dispreference (see main text for details). Experiment 3 (right) combined manipulations. When expressing e.g. preference for an object on a table with two objects (panel 3 from top), the respective object was temporarily enlarged. Condition for Experiment 3 shown here: preference - same speaker - incongruent."}
img <- png::readPNG("figs/method_figure.png")
grid::grid.raster(img)
```

@@

A total of 121 individuals participated in the experiment. The test situation was the same as in the test condition in Experiment 1 (see Fig. \ref{fig:design}, right): One table with object of type A and the other with an object of type A and B. Again, the animal always turned to the table with two objects and ambiguously requested an object. In Experiment 1, the listener had no prior information about each object. In Experiment 3, however, we manipulated common ground expectations in the same way as in Experiment 2. For example, the animal would turn to the table with one object and express that they don’t like object A, then turn to the other table and express that they like object B. Next, after quickly disappearing, they would reappear, turn to the table with two objects and make a request.

For each common ground condition, there were 4 conditions in Experiment 3 resulting from the cross of congruent/incongruent informativeness with same/different speaker. If the preferred/novel object was the less frequent one (object B), the two expectations were congruent. If the preferred/novel object was the more frequent one (object A), expectations were incongruent. For each type of expectation alignment, we varied if the same or a different animal returned. Participants either completed the preference or novelty version with two test trials in each of the four conditions. Before discussing the empirical results, we briefly discuss the model we used to predict expectation integration.

@@

To derive predictions, we used a probabilistic RSA model that simulates a pragmatic listener reasoning about a cooperative speaker who is trying to refer to an object [@frank2012predicting]. The speaker chooses how to refer to the object by reasoning about a naive listener who does not know the labels for the object [@frank2014inferring]. The conditional probability that the listener chooses a referent given an utterance is defined as follows:
$$P_L(r_s|u)\propto P_S(u|r_s)P_S(r_s)$$
Here, $P_S(u|r_s)$ is the likelihood that the speaker will use an utterance $u$ to refer to a specific referent r. It is defined in terms of a utility function $U_S(u;s)$ consisting of the surprisal of $u$ for a naive listener $L_0$, who interprets $u$ according its literal semantics:
$$P_S(u|r_s)\propto exp(\alpha U_S(u;s))$$
The numerical strength of the expression above depends on a scalar value, $\alpha$, which can be interpreted as an indicator of how rational the speaker is in choosing utterances (i.e. as $\alpha$ increases, the speaker is more likely to choose the most informative utterance).

@@

Results are discussed in the form of the proportion with which listeners chose the more informative object (i.e., the object that would be the more informative referent when only considering speaker general expectations). For a comparison to chance within each unique condition see Fig. \ref{fig:plotexp3}. Combinations of alignment and speaker identity differed in how they influenced participants’ responses (GLMM model term: `alignment*speaker`; $\beta$ = `r lm_ex3_r%>%filter(term == "SpeakerSame speaker:AlignmentIncongruent")%>%pull(estimate)`, se = `r lm_ex3_r%>%filter(term == "SpeakerSame speaker:AlignmentIncongruent")%>%pull(std.error)`, *p* `r lm_ex3_r%>%filter(term == "SpeakerSame speaker:AlignmentIncongruent")%>%pull(p.value)`). Fig. \ref{fig:plotmodelcomp}A shows the mean response in each unique condition compared to the pragmatics model. Model predictions and data were highly correlated (r = `r model_comp_r$estimate`, *p* `r model_comp_r$p.value`). Model fit was much better in the model taking into account both types of expectations compared to the uniform prior (BF = `r format(bf_ex3%>%filter(Comparison == "Pragmatic vs. No prior") %>%pull(BayesFactor),digits = 2)`) or prior only model (BF = `r format(bf_ex3%>%filter(Comparison == "Pragmatic vs. Prior only") %>%pull(BayesFactor), digits = 2)`). The inferred noise level in the pragmatics model was 0.27 (95% HDI: 0.21 - 0.34).

@@

As noted above, the strong prior condition was a direct replication of Experiment 3. Results from the two rounds of data collection were highly correlated (r = `r repl_r$estimate`, *p* `r repl_r$p.value`, see Fig. \ref{fig:plotmodelcomp}C). Across levels of prior manipulation, the data from Experiment 4 were highly correlated to the corresponding model predictions (r = `r ex4_comp_r$estimate`, *p* `r ex4_comp_r$p.value`, see Fig. \ref{fig:plotmodelcomp}B). Again, the pragmatics model provided a much better fit compared to the flat prior (BF = `r format(bf_ex4%>%filter(Comparison == "Pragmatic vs. No prior") %>%pull(BayesFactor),digits = 2)`) or prior only model (BF = `r format(bf_ex4%>%filter(Comparison == "Pragmatic vs. Prior only") %>%pull(BayesFactor), digits = 2)`). The inferred noise level in the pragmatics model was 0.28 (95% HDI: 0.24 - 0.32).

@@

Language use and learning requires balancing different types of expectations about one’s interlocutor - expectations about how speakers behave in general and expectations about how a particular speaker might behave in a particular context. Here we used a Bayesian pragmatics model to predict this integration process. Experiment 1 and 2 replicated previous studies showing that adult listeners expect speakers to produce utterances informatively and also with respect to common ground. We then combined the procedures from the first two experiments to study how listeners would integrate expectations. We used the results from Experiment 1 and 2 to specify model parameters that represented the two types of expectations, generating predictions about new behavior. Experiments 3 and 4 showed that both types of expectations influenced listeners inferences. Overall, listener behavior was accurately described by our model, suggesting that listeners trade-off flexibly between speaker specific and general pragmatic expectations.

### 2019-02-05 papers/cogsci2019/cogsci2019.tex
update cogsci submission

Within RSA models, each agent in this recursion is modeled as a Bayesian
reasoner; thus, information integration is treated as a process of
probabilistic inference. The speaker-general informativeness expectation
is already encoded in the structure of the model: Speakers produce
utterances to aid the listener in disambiguating referents. We
operationalize speaker-specific, common ground information as the shared
prior probability of referents in the context of the utterance. Thus, a
natural locus for information integration within these models is the
trade off between the prior probability of a particular referent and the
likelihood of that referent given the current utterance.

Here we evaluate this rational, pragmatic account of information
integration. We isolate speaker-specific and common-ground information
experimentally, then test how adult listeners combine them in a word
learning setting. In Experiments 1 and 2, we replicate findings showing
that listeners expect speakers to a) produce informative utterances
(Experiment 1) and b) communicate about things that are relevant to
common ground (Experiment 2). Based on these results, we generate model
predictions using the RSA framework about how these two components
should be integrated. In Experiment 3, we test how listeners integrate
their expectations and compare model predictions to empirical data.
Experiment 4 replicates and extends Experiment 3 by varying the strength
of common ground assumptions.

@@

For all experiments, we pre-registered the sample size, experimental
design and the statistical analysis. For Experiment 3 and 4, we also
registered the model structure and predictions (see {[}masked for peer
review{]})

@@

\caption[Schematic experimental procedure]{Schematic experimental procedure. In all conditions, at test (bottom), the speaker ambiguously requested an object using a non-word (e.g. “dax”). Participants clicked on the object they thought the speaker referred to. Informativeness (Experiment 1, left) translated to making one object less frequent in context. Common ground (Experiment 2, middle) was manipulated by making one object prefered by or new to the speaker. Green plus signs represent an expression of preference and red minus of dispreference (see main text for details). When expressing e.g. preference for an object on a table with two objects (panel 3 from top), the respective object was temporarily enlarged. Experiment 3 (right) combined manipulations. Condition shown here: preference - same speaker - incongruent.}\label{fig:design}
\end{figure*}
\end{CodeChunk}

@@

A total of 121 individuals participated in the experiment. The test
situation was the same as in the test condition in Experiment 1: One
table with object of type A and the other with an object of type A and
B. Again, the animal always turned to the table with two objects and
ambiguously requested an object. In Experiment 1, the listener had no
prior information about each object. In Experiment 3, however, we
manipulated common ground expectations in the same way as in Experiment
2. For example, the animal would turn to the table with one object and
express that they don't like object A, then turn to the other table and
express that they like object B. Next, after quickly disappearing, they
would reappear, turn to the table with two objects and make a request
(see Fig. \ref{fig:design}, right).

For each common ground condition, there were 4 conditions in Experiment
3 resulting from the cross of congruent/incongruent informativeness with
same/different speaker. If the preferred/novel object was the less
frequent one (object B), the two expectations were congruent. If the
preferred/novel object was the more frequent one (object A),
expectations were incongruent. For each type of expectation alignment,
we varied if the same or a different animal returned. Participants
either completed the preference or novelty version with two test trials
in each of the four conditions. Before discussing the empirical results,
we briefly discuss the cognitive model we used to predict expectation
integration.

@@

To derive predictions, we used a probabilistic RSA model that simulates
a pragmatic listener reasoning about a cooperative speaker who is trying
to refer to an object (Frank \& Goodman, 2012). The speaker chooses how
to refer to the object by reasoning about a naive listener who does not
know the labels for the object (Frank \& Goodman, 2014). The conditional
probability that the listener chooses a referent given an utterance is
defined as follows: \[P_L(r_s|u)\propto P_S(u|r_s)P_S(r_s)\] Here,
\(P_S(u|r_s)\) is the likelihood that the speaker will use an utterance
u to refer to a specific referent r. It is defined in terms of a utility
function \(U_S(u;s)\) consisting of the surprisal of \(u\) for a naive
listener \(L_0\), who interprets \(u\) according its literal semantics:
\[P_S(u|r_s)\propto exp(\alpha U_S(u;s))\] The numerical strength of the
expression above depends on a scalar value, \(\alpha\), which can be
interpreted as an indicator of how rational the speaker is in choosing
utterances (i.e.~as \(\alpha\) increases, the speaker is more likely to
choose the most informative utterance).

@@

Results are discussed in the form of the proportion with which listeners
chose the more informative object (i.e., the object that would be the
more informative referent when only considering speaker general
information). For a comparison to chance within each unique condition
see \ref{fig:plotexp3}. Combinations of alignment and speaker identity
differed in how they influenced participants' responses (GLMM model
term: \texttt{alignment*speaker}; \(\beta\) = -2.92, se = 0.57, \emph{p}
\textless{} .001). Fig. \ref{fig:plotmodelcomp}A and B shows the mean
response in each unique condition compared to the pragmatics model.
Model predictions and data were highly correlated (r = 0.96, \emph{p}
\textless{} .001).Model fit was much better in the model taking into
account both types of expectations compared to the uniform prior (BF =
2e+79) or prior only model (BF = 1.8e+34). The inferred noise level in
the pragmatics model was 0.27 (95\% HDI: 0.21 - 0.34{]}.

@@

As noted above, the strong prior condition was a direct replication of
Experiment 3. Results from the two rounds of data collection were highly
correlated (r = 0.97, \emph{p} \textless{} .001, see Fig.
\ref{fig:plotmodelcomp}C). Across levels of prior manipulation, the data
from Experiment 4 were highly correlated to the corresponding model
predictions (r = 0.91, \emph{p} \textless{} .001, see Fig.
\ref{fig:plotmodelcomp}B). Again, the pragmatics model provided a much
better fit compared to the flat prior (BF = 4.4e+74) or prior only model
(BF = 1.8e+84). The inferred noise level in the pragmatics model was
0.28 (95\% HDI: 0.24 - 0.32{]}.

@@

Language use and learning requires balancing different types of
expectations about one's interlocutor - both expectations about how
speakers behave in general and specific expectations about how a
particular speaker might behave in a particular context. Here we used a
Bayesian pragmatics model to predict this integration process.
Experiment 1 and 2 replicated previous studies showing that adult
listeners expect speakers to produce utterances informatively and also
with respect to common ground. We then combined the procedures from the
first two experiments to study how listeners would integrate
expectations. We used the results from Experiment 1 and 2 to specify
model parameters that represented the two types of expectations,
generating predictions about new behavior. Experiments 3 and 4 showed
that both types of expectations influenced listeners inferences.
Overall, listener behavior was accurately described by our model,
suggesting that listeners trade-off flexibly between speaker specific
and general pragmatic expectations.

@@

\hypertarget{ref-clark2009first}{}
Clark, E. V. (2009). \emph{First language acquisition}. Cambridge
University Press.

\hypertarget{ref-clark1996using}{}
Clark, H. H. (1996). \emph{Using language}. Cambridge University Press.

@@

\hypertarget{ref-grice1991studies}{}
Grice, H. P. (1991). \emph{Studies in the way of words}. Harvard
University Press.

\hypertarget{ref-levinson2000presumptive}{}
Levinson, S. C. (2000). \emph{Presumptive meanings: The theory of
generalized conversational implicature}. MIT press.

@@

\hypertarget{ref-sperber2001relevance}{}
Sperber, D., \& Wilson, D. (2001). \emph{Relevance: Communication and
cognition} (2nd ed.). Oxford; Cambridge, MA: Blackwell Publishers.

\hypertarget{ref-tomasello2009constructing}{}
Tomasello, M. (2009). \emph{Constructing a language}. Harvard university
press.

>>>>

Within RSA models, each agent in the recursion is modeled as a Bayesian
reasoner; thus, information integration is treated as a process of
probabilistic inference. The speaker-general informativeness expectation
is already encoded in the structure of the model: Speakers produce
utterances to aid the listener in disambiguating referents. We
operationalize speaker-specific, common ground information as the shared
prior probability of referents in the context of the utterance. Thus, a
natural locus for information integration within these models is the
trade off between the prior probability of a particular referent and the
likelihood of that referent given the current utterance.

Here we evaluate this rational, pragmatic account of information
integration. We isolate speaker-specific and common-ground information
experimentally, then test how adult listeners combine them in a word
learning setting. In Experiments 1 and 2, we replicate findings showing
that listeners expect speakers to a) produce informative utterances
(Experiment 1) and b) communicate about things that are relevant to
common ground (Experiment 2). Based on these results, we generate model
predictions using the RSA framework about how these two components
should be integrated. In Experiment 3, we test how listeners integrate
their expectations and compare model predictions to empirical data.
Experiment 4 replicates and extends Experiment 3 by varying the strength
of common ground assumptions. For all experiments, we pre-registered the
sample size, experimental design and the statistical analysis. For
Experiment 3 and 4, we also registered the model structure and
predictions (see {[}masked for peer review{]})

@@

@@

\caption[Schematic experimental procedure]{Schematic experimental procedure. In all conditions, at test (bottom), the speaker ambiguously requested an object using a non-word (e.g. “dax”). Participants clicked on the object they thought the speaker referred to. Informativeness (Experiment 1, left) translated to making one object less frequent in context. Common ground (Experiment 2, middle) was manipulated by making one object prefered by or new to the speaker. Green plus signs represent utterances that expressed preference and red minus of dispreference (see main text for details). Experiment 3 (right) combined manipulations. When expressing e.g. preference for an object on a table with two objects (panel 3 from top), the respective object was temporarily enlarged. Condition for Experiment 3 shown here: preference - same speaker - incongruent.}\label{fig:design}
\end{figure*}
\end{CodeChunk}

@@

A total of 121 individuals participated in the experiment. The test
situation was the same as in the test condition in Experiment 1 (see
Fig. \ref{fig:design}, right): One table with object of type A and the
other with an object of type A and B. Again, the animal always turned to
the table with two objects and ambiguously requested an object. In
Experiment 1, the listener had no prior information about each object.
In Experiment 3, however, we manipulated common ground expectations in
the same way as in Experiment 2. For example, the animal would turn to
the table with one object and express that they don't like object A,
then turn to the other table and express that they like object B. Next,
after quickly disappearing, they would reappear, turn to the table with
two objects and make a request.

For each common ground condition, there were 4 conditions in Experiment
3 resulting from the cross of congruent/incongruent informativeness with
same/different speaker. If the preferred/novel object was the less
frequent one (object B), the two expectations were congruent. If the
preferred/novel object was the more frequent one (object A),
expectations were incongruent. For each type of expectation alignment,
we varied if the same or a different animal returned. Participants
either completed the preference or novelty version with two test trials
in each of the four conditions. Before discussing the empirical results,
we briefly discuss the model we used to predict expectation integration.

@@

To derive predictions, we used a probabilistic RSA model that simulates
a pragmatic listener reasoning about a cooperative speaker who is trying
to refer to an object (Frank \& Goodman, 2012). The speaker chooses how
to refer to the object by reasoning about a naive listener who does not
know the labels for the object (Frank \& Goodman, 2014). The conditional
probability that the listener chooses a referent given an utterance is
defined as follows: \[P_L(r_s|u)\propto P_S(u|r_s)P_S(r_s)\] Here,
\(P_S(u|r_s)\) is the likelihood that the speaker will use an utterance
\(u\) to refer to a specific referent r. It is defined in terms of a
utility function \(U_S(u;s)\) consisting of the surprisal of \(u\) for a
naive listener \(L_0\), who interprets \(u\) according its literal
semantics: \[P_S(u|r_s)\propto exp(\alpha U_S(u;s))\] The numerical
strength of the expression above depends on a scalar value, \(\alpha\),
which can be interpreted as an indicator of how rational the speaker is
in choosing utterances (i.e.~as \(\alpha\) increases, the speaker is
more likely to choose the most informative utterance).

@@

Results are discussed in the form of the proportion with which listeners
chose the more informative object (i.e., the object that would be the
more informative referent when only considering speaker general
expectations). For a comparison to chance within each unique condition
see Fig. \ref{fig:plotexp3}. Combinations of alignment and speaker
identity differed in how they influenced participants' responses (GLMM
model term: \texttt{alignment*speaker}; \(\beta\) = -2.92, se = 0.57,
\emph{p} \textless{} .001). Fig. \ref{fig:plotmodelcomp}A shows the mean
response in each unique condition compared to the pragmatics model.
Model predictions and data were highly correlated (r = 0.96, \emph{p}
\textless{} .001). Model fit was much better in the model taking into
account both types of expectations compared to the uniform prior (BF =
2e+79) or prior only model (BF = 1.8e+34). The inferred noise level in
the pragmatics model was 0.27 (95\% HDI: 0.21 - 0.34).

@@

As noted above, the strong prior condition was a direct replication of
Experiment 3. Results from the two rounds of data collection were highly
correlated (r = 0.97, \emph{p} \textless{} .001, see Fig.
\ref{fig:plotmodelcomp}C). Across levels of prior manipulation, the data
from Experiment 4 were highly correlated to the corresponding model
predictions (r = 0.91, \emph{p} \textless{} .001, see Fig.
\ref{fig:plotmodelcomp}B). Again, the pragmatics model provided a much
better fit compared to the flat prior (BF = 4.4e+74) or prior only model
(BF = 1.8e+84). The inferred noise level in the pragmatics model was
0.28 (95\% HDI: 0.24 - 0.32).

@@

Language use and learning requires balancing different types of
expectations about one's interlocutor - expectations about how speakers
behave in general and expectations about how a particular speaker might
behave in a particular context. Here we used a Bayesian pragmatics model
to predict this integration process. Experiment 1 and 2 replicated
previous studies showing that adult listeners expect speakers to produce
utterances informatively and also with respect to common ground. We then
combined the procedures from the first two experiments to study how
listeners would integrate expectations. We used the results from
Experiment 1 and 2 to specify model parameters that represented the two
types of expectations, generating predictions about new behavior.
Experiments 3 and 4 showed that both types of expectations influenced
listeners inferences. Overall, listener behavior was accurately
described by our model, suggesting that listeners trade-off flexibly
between speaker specific and general pragmatic expectations.

@@

\hypertarget{ref-clark2009first}{}
Clark, E. V. (2009). \emph{First language acquisition}. Cambridge:
Cambridge University Press.

\hypertarget{ref-clark1996using}{}
Clark, H. H. (1996). \emph{Using language}. Cambridge: Cambridge
University Press.

@@

\hypertarget{ref-grice1991studies}{}
Grice, H. P. (1991). \emph{Studies in the way of words}. Cambridge, MA:
Harvard University Press.

\hypertarget{ref-levinson2000presumptive}{}
Levinson, S. C. (2000). \emph{Presumptive meanings: The theory of
generalized conversational implicature}. Cambridge, MA: MIT press.

@@

\hypertarget{ref-sperber2001relevance}{}
Sperber, D., \& Wilson, D. (2001). \emph{Relevance: Communication and
cognition} (2nd ed.). Cambridge, MA: Blackwell Publishers.

\hypertarget{ref-tomasello2009constructing}{}
Tomasello, M. (2009). \emph{Constructing a language}. Cambridge, MA:
Harvard University Press.
